Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-06-27 paper_jasa.tex
finished cleaning up power comparison sub-section

\green{Would it make sense to expand this table now to show number of lineups for different experimental conditions?? Maybe not}
\begin{table}[hbtp]
\caption{Number of subjects, gender and total lineups for all three experimental data sets. Note that in some of the lines the number of male and female participants does not add up to the total number of participants due to missing demographic information. }
\begin{center}
\begin{tabular}{ccrrcc}
   \hline 
 Experiment & Subject & Male & Female & Total &Lineup\\ 
    \hline
  1 & 239 & 121 & 107 & 2317 & 60 \\ 
  2 & 351 & 185 & 164 & 3858 & 70 \\ 
  3 & 156 & 106 &  50 & 1681 & 30 \\ 
   \hline
\end{tabular}
\end{center}
\label{tbl:summary}
\end{table}

@@

For each parameter combination, the effect of the lineup is computed as $$E=\beta*\sqrt {n}/\sigma.$$ The experimental data is displayed in figure \ref{fig:power_loess_effect} for different effect sizes. The loess curves fitted to the response data provide the estimated observed power for each test. The corresponding power curves of conventional tests are shown for comparison. The observed power for experiment 2 (scatter plot) is better than what we observed for experiment 1 (box plot). For experiment 3, observed power is always above the conventional test power since for contaminated data we can visually see a trend while conventional test fail to address that.

Figure \ref{fig:power_loess_effect} also reveals that at effect $E=0$, the power is close to 0.05 (Type-I error) for both experiment 1 and 2. For experiment 3 the scenario is different since contamination in the data is visible and this sets the lower limit of proportion correct at a higher value. Thus for fitting model \ref{mixed} to the data we set offset = $log(0.05/0.95)$ so that the estimated power has a fixed lower limit = 0.05 (Type-I error) for effect $E=0$. While fitting model \ref{mixed} we set $W_{\ell i} = E_{\ell i}$, the effect of lineup $\ell$ that subject $i$ has evaluated. Also $E$ is used as the only fixed effect covariate without intercept but offset for experiment 1 and 2. For experiment 3 we fit the intercept to allow the power curves to go above 0.05 as we see in figure \ref{fig:power_loess_effect}.

The power curves obtained by fitting model \ref{mixed} to the data for all the experiments applying different screening criteria is shown in figure \ref{fig:power_screening}. This shows how the screening criteria may affect the results. The corresponding power curves of conventional test are also shown. Notice that no mater what criteria we apply the result does not change much for experiment 2. Also, estimated power curve with criteria 5 for experiment 3 is above the conventional test power curve.

@@

\subsection{Estimating $p$-value in Non-Classical Settings}

>>>>

\begin{table}[hbtp]
\caption{Number of subjects, gender and total lineups seen, and number of distinct lineups for all three experimental data sets. Note that in some of the lines the number of male and female participants does not add up to the total number of participants due to missing demographic information. }
\begin{center}
\begin{tabular}{ccrrcc}
   \hline 
 Experiment & Subject & Male & Female & Total &Lineup\\ 
    \hline
  1 & 239 & 121 & 107 & 2317 & 60 \\ 
  2 & 351 & 185 & 164 & 3858 & 70 \\ 
  3 & 156 & 106 &  50 & 1681 & 30 \\ 
   \hline
\end{tabular}
\end{center}
\label{tbl:summary}
\end{table}

@@

For each parameter combination, the effect is computed as

\[
E=\beta*\sqrt {n}/\sigma.
\]

\noindent Figure \ref{fig:power_loess_effect} shows the power against effect for the three experiments. Data values are represented as dots with size indicating count at this value. The loess curves fit the data, estimating the observed power for different effect sizes, with grey bands indicating simultaneous bootstrap confidence bands. For comparison, the dashed lines show the corresponding power curves of the conventional tests. The observed power for experiments 1 and 2 are lower than the power of the conventional test. This is not a surprise -- it is not expected that visual inference will beat conventional tests, in the scenario when assumptions are satisfied. It is encouraging to see that visual inference mirrors the power vs effect relationship of conventional testing, and comes quite close in experiment 2, where scatterplots are used as the test statistic. Results for experiment 3, are quite different. This is the situation where we expect to see the potential of visual inference, and indeed we do: the power of visual inference is always high, and much higher than the conventional test at small effect sizes.

%(scatter plot) is better than what we observed for experiment 1 (box plot). For experiment 3, observed power is always above the conventional test power since for contaminated data we can visually see a trend while conventional test fail to address that.

It can also be seen that at effect $E=0$, the power is close to 0.05 (Type-I error) for both experiments 1 and 2. For experiment 3 the scenario is different. The contamination in the data is visible which sets the lower limit of proportion correct at a higher value. For this reason, when the subject-specific models (Equation \ref{mixed}) are fit to the data, in the next section, we set offset = $log(0.05/0.95)$ so that the estimated power has a fixed lower limit at 0.05 (Type-I error) when $E=0$. While fitting models (Equation \ref{mixed}) we set $W_{\ell i} = E_{\ell i}$, the effect of lineup $\ell$ that subject $i$ has evaluated. Also $E$ is used as the only fixed effect covariate without intercept but offset for experiment 1 and 2. For experiment 3 an intercept is estimated to allow the power curves to go above 0.05.

%The power curves obtained by fitting model \ref{mixed} to the data for all the experiments applying different screening criteria is shown in figure \ref{fig:power_screening}. This shows how the screening criteria may affect the results. The corresponding power curves of conventional test are also shown. Notice that no mater what criteria we apply the result does not change much for experiment 2. Also, estimated power curve with criteria 5 for experiment 3 is above the conventional test power curve.

@@

\subsection{Estimating $p$-value in non-classical settings}
